\singlespacing A closer look at fixed-effects regression in structural equation modeling using \texttt{lavaan}

### 2020-06-16 article.Rmd
Rerun and generate the main article
Major revisions to text

---
title: "A Closer Look at Fixed-Effects Regression in Structural Equation Modeling Using lavaan"
author: "Henrik Kenneth Andersen"
abstract: "This article provides an in-depth look at fixed-effects regression in the structural equation modeling (SEM) framework, specifically the application of fixed-effects in the `lavaan` package for `R`. It is meant as a applied guide for researchers, covering the underlying model specification, syntax, and summary output. It further discusses various common extentsions to the basic fixed-effect model, demonstrating how to relax model assumptions, deal with measurement error in both the dependent and independent variables, and include time-invariant predictors in a type of fixed-/random-effects hybrid model. \\par \\par \\textbf{Keywords:} Fixed-effects, structural equation modeling, lavaan, R, panel analysis" 
date: "`r Sys.Date()`"
output: 
  pdf_document:
    fig_caption: yes
    number_sections: yes
    includes: 
      in_header: header.tex
documentclass: article
classoption: a4paper
bibliography: references.bib   
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# ----- Packages

library( formatR)
library( knitr)
```

```{r, code=readLines("scripthooks.R"), echo=FALSE, results=FALSE}
```

@@

Several years ago, @Curran2011 reflected positively on the growing use of panel studies in empirical social research. Some of the strengths of panel data are well-known, e.g., the ability to establish temporal precedence, increased statistical power and the reduction of potential alternative models. However, the perhaps greatest strength of panel data is that they allow for a more rigorous testing of substantive theories as a tool for \textit{causal analysis} [@Greene2012]. Panel data, i.e., repeated measures of the same observed units (people, schools, firms, countries, etc.), allow researchers to decompose the error term into a part that stays constant within units and the part that changes over time. This is the essence of fixed-effects (FE) regression. It drastically reduces the number of potential confounders of the relationship between variables of interest by allowing one to control for all time-invariant influences. Thus, by using each unit at an earlier point in time as its own control, and by examining solely \textit{within-unit changes}, we come closer to the \textit{counterfactual} or \textit{potential outcomes} ideal: what would the outcome have been, had the observational unit not been exposed to the treatment [@Angrist2009]?

@@

This article is intended as a guide for researchers looking for in-depth help with specifying FE models in SEM. It focuses on the `lavaan` [@R-lavaan] package for `R` [@R-base]. While `Mplus` [@Mplus] is unarguably the most robust SEM software currently available (in terms of features like alignment, latent variable interactions, for example) , the `lavaan` package has many benefits. First, it and `R` are open source and completely free. For researchers dipping their toes into SEM, there is no financial barrier to try, and no risk if they decide it is not for them. Second, the implementation of `lavaan` in the larger `R` environment is an enormous advantage. Instead of poring over reams of plain text, copying out coefficients by hand, every part of the `lavaan` output is available as an object. This means that all aspects of the fitted `lavaan` object, from fit indices, to coefficients and standard errors, to the model matrices can be accessed and easily integrated into tables and plots. Furthermore, `R` can be used for a great deal of applications. It can be used to manage and manipulate as well as simulate data, perform symbolic algebra, run more traditional analyses (e.g., multiple regression, logistic regression, principal component analysis), etc. Once one is comfortable using `R`, there is no longer any need to switch between different software for data preparation and analysis.

The following article outlines the basic idea of FE regression, the particularities of FE in SEM, and shows its implementation in `lavaan`. Using simulated data (found in the Appendix), it demonstrates and annotates the code for the most basic FE model and provides an overview of the summary output. After that, a number of potential extensions are discussed and demonstrated, from loosening assumptions, to dealing with measurement error, to specifying hybrid fixed- and random effects models in order to include time-invariant predictors.

@@

The most basic fixed-effects model can be expressed as
\begin{align}
y_{it} & = \bm{x_{it}}\bm{\beta} + \alpha_{i} + \varepsilon_{it}, \label{fe}
\end{align}
where $y_{it}$ is the observed outcome variable for unit $i$ at time $t$, $\bm{x_{it}}$ is a $1 \times k$ vector of covariates linked to the outcome by the $k \times 1$ vector of coefficients $\bm{\beta}$. All stable, unit-specific characteristics such as date of birth, country of origin, sex, etc. are captured by the time-constant part of the error $\alpha_{i}$, whereas $\varepsilon_{it}$ is the idiosyncratic error term that varies between units and over time [@Bruederl2015].

The point of FE regression is to be able to drop assumptions regarding the relatedness of the model covariates and the unobserved time-constant effects, i.e., $\E[\bm{x_{it}}^{\intercal}\alpha_{i}]$. We can do so by manipulating Equation \eqref{fe} so that the individual effect, $\alpha_{i}$ is eliminated. In doing so, we no longer need to worry about potential bias due to unit-specific characteristics. This can be achieved in a number of ways (e.g., differencing, least squares dummy variable regression), but the most common method involves subtracting the over-time unit mean from each of the terms: 
\begin{align}
(y_{it} - \bar{y}_{i}) & = (\bm{x_{it}} - \bar{\bm{x}_{i}})\bm{\beta} + (\alpha_{i} - \bar{\alpha}_{i}) + (\varepsilon_{it} - \bar{\varepsilon}_{i}) \\
\ddot{y}_{it} & = \bm{\ddot{x}_{it}}\bm{\beta} + \ddot{\varepsilon}_{it}.
\end{align}
Because the average of something that does not change is that thing itself, the individual effects get wiped out by the demeaning. This means that no assumptions about the relatedness of the model covariates and the unit-specific portion of the error are needed. The unbiasedness of the estimate is related solely to the strict exogeneity assumption imposed on the indiosyncratic errors, i.e., $\E[\ddot{\varepsilon}_{it}|\bm{\ddot{x}_{it}}] = 0$ which also implies $\E[\bm{\ddot{x}}_{is}^{\intercal}\ddot{\varepsilon}_{it}] = \bm{0}, \ \forall \ s, t = 1, ..., T$ [@Bruederl2015; @Wooldridge2002]. Strict exogeneity not only rules out contemporaneous correlations between the covariates and the error, but it also implies that for the estimates $\bm{\beta}$ to be unbiased, the covariates at all timepoints must not be correlated with the error at all timepoints.

@@

Analytically, Equation \eqref{fe} applies as well to FE models in SEM as we will see shortly. However, there are a number of differences we must address. First, SEM is a covariance-based approach. This means that the vectors of individual observations are only of interest insomuch as they can be converted into observed covariances. Second, the way in which we use a latent variable to represent the individual effects forces us to convert the data from stacked, long-format vectors of length $NT$ to $T$ wide-format vectors of length $N$. Lastly, the notation in SEM differs from traditional least squares-based models. There are a number of different model notations (see, for example @Bollen1989 for an overview), but the one that will serve us best is one that was proposed by @Graff1979:
\begin{align}
\bm{y}^{+} & = \bm{\Lambda_{y}}^{+} \bm{\eta}^{+}, \\
\bm{\eta}^{+} & = \bm{\eta}^{+}\bm{B} + \bm{\zeta}^{+}, 
\end{align}
where $\bm{\eta}^{+} = (\bm{y}, \bm{x}, \bm{\eta}, \bm{\xi})^{\intercal}$, $\bm{\zeta}^{+} = (\bm{\varepsilon}, \bm{\delta}, \bm{\zeta}, \bm{\xi})^{\intercal}$, $\bm{y}^{+} = (\bm{y}, \bm{x})^{\intercal}$. Furthermore, $\bm{y}$ is a vector of observed dependent variables and $\bm{x}$ is a vector of observed independent variables. $\bm{\eta}$ is a vector of the latent dependent variables and $\bm{\xi}$ is a vector of latent independent variables. $\bm{\varepsilon}$ and $\bm{\delta}$ are vectors of the errors of the observed dependent and independent variables, respectively, and $\bm{\zeta}$ is a vector of the errors, or disturbances, of the latent variables. Notice the $^{+}$ symbol is just meant to differentiate the vectors with them from those without them. That means, $\bm{\eta}^{+}$ is a vector that holds the observed and latent variables, both dependent (i.e., 'endogenous') and independent (i.e., 'exogeneous'), $\bm{\zeta}^{+}$ holds the errors for the observed variables and the disturbances of the latent variables. $\bm{y}^{+}$ holds just the observed variables, both dependent and independent, and $\bm{\Lambda_{y}}^{+}$ is a matrix of ones and zeros that selects the observed variables from $\bm{\eta}^{+}$. Lastly $\bm{B}$ is a matrix that holds the regression coefficients.

@@

@@

This notation may be confusing at first, but it has advantages. First, it allows us to be more flexible with our models. For example, it allows observed $\bm{x}$ to influence observed $\bm{y}$ directly (more common notation assumes that substantive effects occur only between latent variables, observed ones are only used as indicators). It also allows $\bm{\xi}$, i.e., any latent exogenous variables, to influence $\bm{y}$ directly. These two scenarios cover the traditional FE model with observed variables, and one in which latent variables are used to account for measurement error in the independent variables, which will be discussed later in this article. It is also consistent with the notation used in popular SEM software like `lavaan` and `Mplus`. That means the matrix representation of the model one sees if they type in `lavInspect( model, what = "est")` after specifying their model in `lavaan` will match up with the notation used here. Let us, however, make things more concrete and take a look at a simple, three-wave version of the typical FE model in SEM using this notation (see Figure \ref{fig:fesem}). For that, we have:
\begin{align}
\begin{split}
\bm{y}^{+} & = \bm{\Lambda_{y}}^{+}\bm{\eta}^{+} \\
\begin{pmatrix}
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3}
\end{pmatrix} & = 
\begin{blockarray}{cccccccc}
 & y_{1} & y_{2} & y_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
 \begin{block}{c(ccccccc)}
 y_{1} & 1 & 0 & 0 & 0 & 0 & 0 & 0 \\
 y_{2} & 0 & 1 & 0 & 0 & 0 & 0 & 0 \\
 y_{3} & 0 & 0 & 1 & 0 & 0 & 0 & 0 \\ 
 x_{1} & 0 & 0 & 0 & 1 & 0 & 0 & 0 \\
 x_{2} & 0 & 0 & 0 & 0 & 1 & 0 & 0 \\
 x_{3} & 0 & 0 & 0 & 0 & 0 & 1 & 0 \\
 \end{block}
\end{blockarray}
\begin{pmatrix}
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3} \\
\alpha
\end{pmatrix},
\end{split}
\end{align}
\begin{align}
\begin{split}
\bm{\eta}^{+} & = \bm{\eta}^{+}\bm{B} + \bm{\zeta}^{+} \\
\begin{pmatrix}
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3} \\
\alpha
\end{pmatrix} & = 
\begin{pmatrix}
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3} \\
\alpha
\end{pmatrix}
\begin{blockarray}{cccccccc}
 & y_{1} & y_{2} & y_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
 \begin{block}{c(ccccccc)}
 y_{1}  & 0 & 0 & 0 & \beta & 0 & 0 & 1 \\
 y_{2}  & 0 & 0 & 0 & 0 & \beta & 0 & 1\\
 y_{3}  & 0 & 0 & 0 & 0 & 0 & \beta & 1 \\ 
 x_{1}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 x_{2}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 x_{3}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 \alpha & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 \end{block}
\end{blockarray} + 
\begin{pmatrix}
\varepsilon_{1} \\
\varepsilon_{2} \\
\varepsilon_{3} \\
x_{1} = \delta_{1} \\
x_{2} = \delta_{2} \\
x_{3} = \delta_{3} \\
\alpha = \xi \\
\end{pmatrix}. \label{sem_matrix}
\end{split}
\end{align}
Notice in $\bm{\zeta}^{+}$, for the independent variables it does not matter whether we write, for example $x_{1}$ or $\delta_{1}$. By treating these variables as exogenous, we can say their entire variance is made up of 'error', i.e., it is the combined effect of all unobserved influences.

Admittedly, Equation \eqref{sem_matrix} may not look like much yet. We can remedy this by first putting the equation for $\bm{\eta}^{+}$ in reduced form, i.e., by getting rid of the dependent variable on the r.h.s.: 
\begin{align}
\bm{\eta}^{+} & = \bm{\eta}^{+}\bm{B} + \bm{\zeta}^{+} \\
\bm{\eta}^{+} - \bm{\eta}^{+}\bm{B} & = \bm{\zeta}^{+} \\
(\bm{I} - \bm{B})\bm{\eta}^{+} & = \bm{\zeta}^{+} \\
\bm{\eta}^{+} & = (\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+},
\end{align}
where $\bm{I}$ is the identity matrix. By substituting this back into the equation for the observed variables we get $\bm{y}^{+} = \bm{\Lambda_{y}}^{+}[(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+}]$, which works out to:
\begin{align}
\begin{pmatrix} 
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3} \\
\end{pmatrix} & = 
\begin{pmatrix}
\alpha + \beta x_{1} + \varepsilon_{1} \\
\alpha + \beta x_{2} + \varepsilon_{2} \\
\alpha + \beta x_{3} + \varepsilon_{3} \\
x_{1} \\
x_{2} \\
x_{3}
\end{pmatrix},
\end{align}
which is of course exactly what we should expect given Equation \eqref{fe}.

>>>>

---
title: \singlespacing A closer look at fixed-effects regression in structural equation modeling using \texttt{lavaan}
author: Henrik Kenneth Andersen
abstract: "This article provides an in-depth look at fixed-effects regression in the structural equation modeling (SEM) framework, specifically the application of fixed-effects in the `lavaan` package for `R`. It is meant as a applied guide for researchers, covering the underlying model specification, syntax, and summary output. Online supplementary materials further discuss various common extentsions to the basic fixed-effect model, demonstrating how to relax model assumptions, deal with measurement error in both the dependent and independent variables, and include time-invariant predictors in a type of hybrid fixed-/random-effects model. \\par  \\textbf{Keywords:} Fixed-effects, structural equation modeling, lavaan, R, panel analysis" 
date: "`r Sys.Date()`"
output: 
  pdf_document:
    fig_caption: yes
    number_sections: yes
    includes: 
      in_header: header.tex
documentclass: article
classoption: a4paper
urlcolor: blue
link-citations: yes
bibliography: references.bib   
---

```{r setup, include=FALSE}
# Knitr options 
knitr::opts_chunk$set(echo = TRUE)

# Packages (some are repeated below, add them here as well just to be sure)
library(formatR)
library(knitr)
library(lavaan)

hook_chunk = knitr::knit_hooks$get('chunk')

knitr::knit_hooks$set(chunk = function(x, options) {
  regular_output = hook_chunk(x, options)
  # add latex commands if chunk option singlespacing is TRUE
  if (isTRUE(options$singlespacing)) 
    sprintf("\\singlespacing\n %s \n\\doublespacing", regular_output)
  else
    regular_output
})

knitr::opts_chunk$set(echo = TRUE, singlespacing = TRUE)

# Script hooks for chunk outputs
source("scripthooks.R")
```

@@

Several years ago, @Curran2011 reflected positively on the growing use of panel studies in empirical social research. Some of the strengths of panel data are well-known, e.g., the ability to establish temporal precedence, increased statistical power and the reduction of potential alternative models. However, perhaps the greatest strength of panel data is that they allow for a more rigorous testing of substantive theories. Panel data, i.e., repeated measures of the same observed units (people, schools, firms, countries, etc.), allow researchers to decompose the error term into a part that stays constant within units and the part that changes over time. The part that does not change over time can be seen as the combined effect of all time-invariant influences (e.g., sex, date of birth, nationality) on the dependent variable. Fixed-effects (FE) regression involves controlling for these time-invariant influences via a number of various methods. It thus drastically reduces the number of potential confounders of the relationship between variables of interest.

@@

This article is intended as a guide for researchers looking for in-depth help with specifying FE models in SEM. It focuses on the `lavaan` [@R-lavaan] package for `R` [@R-base]. While `Mplus` [@Mplus] is arguably the most robust SEM software currently available (in terms of features like alignment, latent variable interactions, for example) , the `lavaan` package has many benefits. First, it and `R` are open source and completely free. For researchers dipping their toes into SEM, there is no financial barrier to try, and no risk if they decide it is not for them. Second, the implementation of `lavaan` in the larger `R` environment is an enormous advantage. Instead of poring over reams of plain text, copying out coefficients by hand, every part of the `lavaan` output is available as an object. This means that all aspects of the model, from fit indices, to coefficients and standard errors, to the model matrices, can be accessed and easily integrated into tables and plots. Furthermore, `R` can be used for a great deal of applications. It can be used to manage and manipulate as well as simulate data, perform symbolic algebra, run more traditional analyses (e.g., multiple regression, logistic regression, principal component analysis), etc. Once one is comfortable using `R`, there is no longer any need to switch between different software for data preparation and analysis.

The following article outlines the basic idea of FE regression, the particularities of FE in SEM, and shows its implementation in `lavaan`. Using [simulated data](https://github.com/henrik-andersen/FE-SEM/blob/master/simulation-code.R), it demonstrates and annotates the code for the most basic FE model and provides an overview of the summary output. A number of potential extensions to the basic model, including relaxing various assumptions, dealing with measurement error in both the independent and dependent variables, as well as the inclusion of time-invariant predictors in the form of a hybrid fixed-/random-effects model, are shown in detail in the form of [online supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/extensions.pdf).

@@

To begin, let us review the conventional, least squares-based FE model. It is normally written as
\begin{align}
y_{it} & = \bm{x_{it}}\bm{\beta} + \alpha_{i} + \varepsilon_{it} \label{eq:fe}, 
\end{align}
where $y_{it}$ is the observed outcome variable for unit $i$ at time $t$, $\bm{x_{it}}$ is a $1 \times k$ vector of covariates linked to the outcome by the $k \times 1$ vector of coefficients $\bm{\beta}$. All stable, unit-specific characteristics such as date of birth, country of origin, sex, etc. are captured by the time-constant part of the error $\alpha_{i}$, whereas $\varepsilon_{it}$ is the idiosyncratic error term that varies between units and over time [@Bruederl2015].

At least a subset of all the possible time-invariant, unit-specific characteristics affecting the dependent variable are unobservable. Some, like sex and nationality, will be available in a typical dataset, but others, like (for example) ability or motivation will remain unobserved. If the conditional mean of any of these unit-specific characteristics depends on the model covariates, i.e., $\E[\alpha_{i}|\bm{x_{it}}] \ne \E[\alpha_{i}]$ (which implies $\Cov(\bm{x_{it}}^{\intercal},\alpha_{i}) \ne \bm{0}$), then the coefficient of interest, $\beta$, will be biased.

The point of FE regression is to be able to drop assumptions regarding the relatedness of the model covariates and the unobserved time-constant effects. We can do so by manipulating Equation \eqref{eq:fe} so that the individual effect, $\alpha_{i}$ is eliminated. In doing so, we no longer need to worry about potential bias due to unit-specific characteristics. This can be achieved in a number of ways (e.g., differencing, least squares dummy variable regression), but the most common method involves subtracting the over-time unit mean from each of the terms: 
\begin{align}
(y_{it} - \bar{y}_{i}) & = (\bm{x_{it}} - \bar{\bm{x}}_{i})\bm{\beta} + (\alpha_{i} - \bar{\alpha}_{i}) + (\varepsilon_{it} - \bar{\varepsilon}_{i}) \\
\ddot{y}_{it} & = \bm{\ddot{x}_{it}}\bm{\beta} + \ddot{\varepsilon}_{it}.
\end{align}
Because the average of something that does not change is that thing itself, the individual effects get wiped out by the demeaning. This means that no assumptions about the relatedness of the model covariates and the unit-specific portion of the error are needed. The unbiasedness of the estimate is related solely to the strict exogeneity assumption imposed on the idiosyncratic errors, i.e., $\E[\ddot{\varepsilon}_{it}|\bm{\ddot{x}_{it}}] = \E[\ddot{\varepsilon}_{it}] = 0$ which also implies $\E[\bm{\ddot{x}}_{is}^{\intercal}\ddot{\varepsilon}_{it}] = \bm{0}, \ \forall \ s, t = 1, ..., T$ [@Bruederl2015; @Wooldridge2002]. The assumption of strict exogeneity comes from the demeaning, where $\E[\bm{\ddot{x}}_{it}^{\intercal}\ddot{\varepsilon}_{it}] = \E[(\bm{x_{it}} - \bm{\bar{x}_{i}})^{\intercal}(\varepsilon_{it} - \bar{\varepsilon}_{i})] = \E[(\bm{x_{it}} - T^{-1}(\bm{x_{i1}} + \bm{x_{i2}} + ... + \bm{x_{iT}}))(\varepsilon_{it} - T^{-1}(\varepsilon_{i1} + \varepsilon_{i2} + ... + \varepsilon_{iT}))]$ which implies that for $\E[\bm{\ddot{x}_{it}}^{\intercal}\ddot{\varepsilon}_{it}]$ to hold, the covariates must be uncorrelated not just with the contemporaneous errors, but the errors at all timepoints.

@@

Analytically, Equation \eqref{eq:fe} applies as well to FE models in SEM as we will see shortly. However, there are a number of differences we must address. First, SEM is a covariance-based approach. This means that the vectors of individual observations are only of interest insomuch as they can be converted into observed covariances. Second, the way in which we use a latent variable to represent the individual effects forces us to convert the data from stacked, long-format vectors of length $NT$ to $T$ individual vectors of length $N$. Lastly, the notation in SEM differs from traditional least squares-based models. There are a number of different model notations (see, for example @Bollen1989 for an overview), but the one that will serve us best is one that was proposed by @Graff1979:
\begin{align}
\bm{y}^{+} & = \bm{\Lambda_{y}}^{+} \bm{\eta}^{+}, \\
\bm{\eta}^{+} & = \bm{\eta}^{+}\bm{B} + \bm{\zeta}^{+}, 
\end{align}
where $\bm{\eta}^{+} = (\bm{y}, \bm{x}, \bm{\eta}, \bm{\xi})^{\intercal}$, $\bm{\zeta}^{+} = (\bm{\varepsilon}, \bm{\delta}, \bm{\zeta}, \bm{\xi})^{\intercal}$, $\bm{y}^{+} = (\bm{y}, \bm{x})^{\intercal}$. $\bm{y}$ is a vector of observed dependent variables and $\bm{x}$ is a vector of observed independent variables. $\bm{\eta}$ is a vector of the latent dependent variables and $\bm{\xi}$ is a vector of latent independent variables. $\bm{\varepsilon}$ and $\bm{\delta}$ are vectors of the errors of the observed dependent and independent variables, respectively, and $\bm{\zeta}$ is a vector of the errors, or disturbances, of the latent variables. Notice the $^{+}$ symbol is just meant to differentiate the vectors with them from those without them. That means, $\bm{\eta}^{+}$ is a vector that holds the observed and latent variables, both dependent (in SEM they are referred to as 'endogenous') and independent (i.e., 'exogeneous')^[For our purposes, the terms endogenous and dependent, on the one hand, and exogenous and independent, on the other, can be used interchangeably.], $\bm{\zeta}^{+}$ holds the errors for the observed variables and the disturbances of the latent variables. $\bm{y}^{+}$ holds just the observed variables, both dependent and independent, and $\bm{\Lambda_{y}}^{+}$ is a matrix of ones and zeros that selects the observed variables from $\bm{\eta}^{+}$. Lastly $\bm{B}$ is a matrix that holds the regression coefficients.

@@

This notation may be confusing at first, but it has advantages. First, it allows us to be more flexible with our models. For example, it allows observed $\bm{x}$ to directly influence observed $\bm{y}$ (more common notation assumes that substantive effects occur only between latent variables, observed ones are only used as indicators, see for example @Bollen1989, @Kline2016). It also allows $\bm{\xi}$, i.e., any latent exogenous variables, to influence $\bm{y}$ directly. These two scenarios cover the traditional FE model with observed variables, and one in which latent variables are used to account for measurement error in the independent variables. It is also consistent with the notation used in popular SEM software like `lavaan` and `Mplus`. That means the matrix representation of the model one sees if they type in `lavInspect(model, what = "est")` after specifying their model in `lavaan` will match up with the notation used here.^[In `Mplus`, the model matrices can be requested by including `OUTPUT: TECH1` in the input file.] Let us, however, make things more concrete and take a look at a simple, three-wave version of the typical FE-SEM using this notation (shown graphically in Figure \ref{fig:fesem}). For that, we have:
\begin{align}
\begin{split}
\bm{y}^{+} & = \bm{\Lambda_{y}}^{+}\bm{\eta}^{+} \\
\begin{pmatrix}
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3}
\end{pmatrix} & = 
\begin{blockarray}{cccccccc}
 & y_{1} & y_{2} & y_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
 \begin{block}{c(ccccccc)}
 y_{1} & 1 & 0 & 0 & 0 & 0 & 0 & 0 \\
 y_{2} & 0 & 1 & 0 & 0 & 0 & 0 & 0 \\
 y_{3} & 0 & 0 & 1 & 0 & 0 & 0 & 0 \\ 
 x_{1} & 0 & 0 & 0 & 1 & 0 & 0 & 0 \\
 x_{2} & 0 & 0 & 0 & 0 & 1 & 0 & 0 \\
 x_{3} & 0 & 0 & 0 & 0 & 0 & 1 & 0 \\
 \end{block}
\end{blockarray}
\begin{pmatrix}
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3} \\
\alpha
\end{pmatrix},
\end{split} \label{eq:y}
\end{align}

\begin{align}
\begin{split}
\bm{\eta}^{+} & = \bm{\eta}^{+}\bm{B} + \bm{\zeta}^{+} \\
\begin{pmatrix}
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3} \\
\alpha
\end{pmatrix} & = 
\begin{pmatrix}
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3} \\
\alpha
\end{pmatrix}
\begin{blockarray}{cccccccc}
 & y_{1} & y_{2} & y_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
 \begin{block}{c(ccccccc)}
 y_{1}  & 0 & 0 & 0 & \beta & 0 & 0 & 1 \\
 y_{2}  & 0 & 0 & 0 & 0 & \beta & 0 & 1\\
 y_{3}  & 0 & 0 & 0 & 0 & 0 & \beta & 1 \\ 
 x_{1}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 x_{2}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 x_{3}  & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 \alpha & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
 \end{block}
\end{blockarray} + 
\begin{pmatrix}
\varepsilon_{1} \\
\varepsilon_{2} \\
\varepsilon_{3} \\
x_{1} = \delta_{1} \\
x_{2} = \delta_{2} \\
x_{3} = \delta_{3} \\
\alpha = \xi \\
\end{pmatrix}. 
\end{split} \label{eq:eta}
\end{align} 
Notice in $\bm{\zeta}^{+}$, for the independent variables it does not matter whether we write, for example $x_{1}$ or $\delta_{1}$. By treating these variables as exogenous, we can say their entire variance is made up of 'error', i.e., it is the combined effect of all unobserved influences.

Admittedly, Equations \eqref{eq:y} and \eqref{eq:eta} may not look like much yet. We can remedy this by first putting the equation for $\bm{\eta}^{+}$ in reduced form, i.e., by getting rid of the dependent variable on the r.h.s.: 
\begin{align}
\bm{\eta}^{+} & = \bm{\eta}^{+}\bm{B} + \bm{\zeta}^{+} \\
\bm{\eta}^{+} - \bm{\eta}^{+}\bm{B} & = \bm{\zeta}^{+} \\
(\bm{I} - \bm{B})\bm{\eta}^{+} & = \bm{\zeta}^{+} \\
\bm{\eta}^{+} & = (\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+},
\end{align}
where $\bm{I}$ is the identity matrix. By substituting this back into the equation for the observed variables we get $\bm{y}^{+} = \bm{\Lambda_{y}}^{+}[(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+}]$, which works out to:
\begin{align}
\begin{pmatrix} 
y_{1} \\
y_{2} \\
y_{3} \\
x_{1} \\
x_{2} \\
x_{3} \\
\end{pmatrix} & = 
\begin{pmatrix}
\alpha + \beta x_{1} + \varepsilon_{1} \\
\alpha + \beta x_{2} + \varepsilon_{2} \\
\alpha + \beta x_{3} + \varepsilon_{3} \\
x_{1} \\
x_{2} \\
x_{3}
\end{pmatrix},
\end{align}
which is of course exactly what we should expect given Equation \eqref{eq:fe}^[We can use the `sympy` package in `python` to verify and show the steps for this and other examples, see the [supplementary materials](https://github.com/henrik-andersen/FE-SEM/blob/master/sympy-doublecheck-matrixnotation.py) for the code.].

@@

What essentially differentiates an FE from a random effects (RE) model is our assumption concerning the relationship between the unobserved individual effects and the model covariates [@Bollen2010]. The FE model assumes that $\E[\alpha x_{t}] \ne 0$. As such, if we fail to control for the correlation of the covariate and the time-invariant part of the error, then the coefficient of interest, here $\beta$, will be biased. Our assumption regarding whether the individual effects are correlated with the model covariates occurs in $\E[\bm{\zeta}^{+}\bm{\zeta}^{+ \intercal}] = \bm{\Psi}$, the covariance matrix of the errors
\begin{align}
\bm{y}^{+}\bm{y}^{+ \intercal} & = \E[(\bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+})(\bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+})^{\intercal}] \\
 & = \E[(\bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1}\bm{\zeta}^{+})(\bm{\zeta}^{+ \intercal}(\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{+ \intercal})] \\
 & = \bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1} \E[\bm{\zeta}^{+}\bm{\zeta}^{+ \intercal}] (\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{+ \intercal} \\
 & = \bm{\Lambda_{y}}^{+}(\bm{I} - \bm{B})^{-1} \bm{\Psi} (\bm{I} - \bm{B})^{-1 \intercal}\bm{\Lambda_{y}}^{+ \intercal}.
\end{align}
In the case of an FE model, $\bm{\Psi}$ will reflect our belief that the individual effects are correlated with the model covariates, here again for demonstration the three-wave model:
\begin{align}
\bm{\Psi} & =
\begin{blockarray}{cccccccc}
 & \varepsilon_{1} & \varepsilon_{2} & \varepsilon_{3} & x_{1} & x_{2} & x_{3} & \alpha \\
 \begin{block}{c(ccccccc)}
 \varepsilon_{1} & \varepsilon_{1}^{2} &                     &                     &              &              &              & \\
 \varepsilon_{2} & 0                   & \varepsilon_{2}^{2} &                     &              &              &              & \\
 \varepsilon_{3} & 0                   & 0                   & \varepsilon_{3}^{2} &              &              &              & \\
 x_{1}           & 0                   & 0                   & 0                   & x_{1}^{2}    &              &              & \\
 x_{2}           & 0                   & 0                   & 0                   & x_{2}x_{1}   & x_{2}^{2}    &              & \\
 x_{3}           & 0                   & 0                   & 0                   & x_{3}x_{1}   & x_{3}x_{2}   & x_{3}^{2}    & \\
 \alpha          & 0                   & 0                   & 0                   & \alpha x_{1} & \alpha x_{2} & \alpha x_{3} & \alpha^{2} \\
 \end{block}
\end{blockarray}.
\end{align}
Knowing this, we can work out the equation for the coefficient of interest, $\beta$. For the sake of simplicity, assume here and throughout mean-centered variables: 
\begin{align}
\Cov(y_{t},x_{t}) & = \E[y_{t}x_{t}] \\
 & = \E[(\alpha + \beta x_{t} + \varepsilon_{t})x_{t}] \\
 & = \E[\alpha x_{t} + \beta x_{t}^{2} + \varepsilon_{t}x_{t}] \\
 & = \Cov(\alpha, x_{t}) + \beta \Var(x_{t}) \\
\hat{\beta} & = \frac{\Cov(y_{t},x_{t}) - \Cov(\alpha, x_{t})}{\Var(x_{t})}. 
\end{align}
This should make intuitive sense. From the observed covariance between the dependent and the independent variable, we are partialling out the part that is due to the covariance between the independent variable and the individual effects per unit, and then dividing by the variance of the independent variable, as usual. For the RE model, we assume $\E[\alpha x_{t}] = 0$ and the equation reduces to $\hat{\beta} = \Cov(y_{t},x_{t})/\Var(x_{t})$. The rest of the model-implied covariance matrix results from $\bm{y}^{+}\bm{y}^{+ \intercal}$.
